Supplementary Materials for cILR: Taxonomic enrichment analysis with isometric log ratios

### 2021-08-20 manuscript_plos.tex

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Research in human associated microbiomes primarily involves the analysis of taxonomic count tables generated via high-throughput sequencing. It is difficult to apply statistical tools as the data is high-dimensional, sparse, and strictly compositional. An approachable way to alleviate high-dimensionality and sparsity is to aggregate variables into pre-defined sets. Set-based analysis is ubiquitous in the genomics literature, and has demonstrable impact in improving interpretability and power of downstream analysis. Unfortunately, there is a lack of interest in developing more sophisticated set-based analysis methods specific to microbiome data, where current practice often employs abundance summation as a technique for aggregation. This approach prevents comparison across sets of different sizes, does not preserve inter-sample distances, and amplifies protocol bias. Here, we attempt to fill this gap with a new single sample taxon set enrichment method based on the isometric log ratio transformation and the competitive null hypothesis commonly used in the enrichment analysis literature. Our approach, titled competitive isometric log ratio (cILR), generates sample-specific enrichment scores as the scaled log ratio of the subcomposition defined by taxa within a set and the subcomposition defined by its complement. We provide sample-level significance testing by estimating an empirical null distribution of our test statistic with valid p-values. Herein we demonstrate under both real data applications and simulations that cILR controls for type I error even under high sparsity and high inter-taxa correlation scenarios, while additionally providing informative scores that can be inputs to downstream differential abundance and prediction tasks.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
The study of human associated microbiomes rely on genomic surveys via high-throughput sequencing. However, microbiome data is sparse and high dimensional which prevents the application of standard statistical techniques. One approach to address this problem is to perform analyses at the level of taxon sets. Set-based analysis has a long history in the genomics literature, with demonstrable impact in improving both power and interpretability. Unfortunately, there is not a lot of research in developing new set-based tools for microbiome data specifically, given that compared to other omics data types microbiome data is strictly compositional. We developed a new tool to generate taxon set enrichment scores at the sample level by combining the isometric log-ratio and the competitive null hypothesis. Our scores can be used for statistical inference, and as inputs to other downstream analyses such as differential abundance and prediction models. We demonstrate the performance of our method against competing approaches across both real data analyses and simulation studies.

@@

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}
The microbiome is the collection of microorganisms (bacteria, protozoa, archaea, fungi, and viruses) which co-exist with their human host. Previous research has shown that changes in the composition of the microbiome are associated with important health outcomes such as inflammatory bowel disease \cite{proctor2019}, type II diabetes \cite{sharma2019}, and obesity \cite{aoun2020}. To understand the central role of the microbiome in human health, researchers have relied on high-throughput sequencing methods, either by targeting a specific representative gene (i.e. amplicon sequencing) or by profiling all the genomic content of the sample (i.e. whole-genome shotgun sequencing) \cite{cho2012}. Processing raw sequencing data using a variety of bioinformatic pipelines \cite{callahan2016,truong2015} yields taxonomic abundance tables that can be used in downstream statistical analyses to identify associations between an outcome/exposure and identified taxa.

However, there exists unique challenges in the analysis of these data tables \cite{li2019,li2015}. First, like other sequencing-based datasets, microbiome count data is often high dimensional, where the number of detected taxa far exceeds the number of samples usually present. For predictive tasks, microbiome-specific penalized regression approaches have been developed to address this issue \cite{shi2016}, allowing for simultaneous model fitting and variable selection. For differential abundance tasks, researchers often utilize multiple hypothesis correction methods \cite{sankaran2014,benjamini1995} or omnibus tests \cite{chen2018} to address hypothesis testing burden.

>>>>

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Research in human associated microbiomes often involves the analysis of taxonomic count tables generated via high-throughput sequencing. It is difficult to apply statistical tools as the data is high-dimensional, sparse, and strictly compositional. An approachable way to alleviate high-dimensionality and sparsity is to aggregate variables into pre-defined sets. Set-based analysis is ubiquitous in the genomics literature, and has demonstrable impact in improving interpretability and power of downstream analysis. Unfortunately, there is a lack of sophisticated set-based analysis methods specific to microbiome data, where current practice often employs abundance summation as a technique for aggregation. This approach prevents comparison across sets of different sizes, does not preserve inter-sample distances, and amplifies protocol bias. Here, we attempt to fill this gap with a new single sample taxon set enrichment method based on the isometric log ratio transformation and the competitive null hypothesis commonly used in the enrichment analysis literature. Our approach, titled competitive isometric log ratio (cILR), generates sample-specific enrichment scores as the scaled log ratio of the subcomposition defined by taxa within a set and the subcomposition defined by its complement. We provide sample-level significance testing by estimating an empirical null distribution of our test statistic with valid p-values. Herein we demonstrate using both real data applications and simulations that cILR controls for type I error even under high sparsity and high inter-taxa correlation scenarios. Additionally, it provides informative scores that can be inputs to downstream differential abundance and prediction tasks.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
The study of human associated microbiomes relies on genomic surveys via high-throughput sequencing. However, microbiome data is sparse and high dimensional which prevents the application of standard statistical techniques. One approach to address this problem is to perform analyses at the level of taxon sets. Set-based analysis has a long history in the genomics literature, with demonstrable impact in improving both power and interpretability. Unfortunately, there is not a lot of research in developing new set-based tools for microbiome data specifically, given that compared to other 'omics data types microbiome data is strictly compositional. We developed a new tool to generate taxon set enrichment scores at the sample level by combining the isometric log-ratio and the competitive null hypothesis. Our scores can be used for statistical inference, and as inputs to other downstream analyses such as differential abundance and prediction models. We demonstrate the performance of our method against competing approaches across both real data analyses and simulation studies.

@@

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}
The microbiome is the collection of microorganisms (bacteria, protozoa, archaea, fungi, and viruses) which co-exist with their host. Previous research has shown that changes in the composition of the human gut microbiome are associated with important health outcomes such as inflammatory bowel disease \cite{proctor2019}, type II diabetes \cite{sharma2019}, and obesity \cite{aoun2020}. To understand the central role of the microbiome in human health, researchers have relied on high-throughput sequencing methods, either by targeting a specific representative gene (i.e. amplicon sequencing) or by profiling all the genomic content of the sample (i.e. whole-genome shotgun sequencing) \cite{cho2012}. Raw sequencing data is then processed through a variety of bioinformatic pipelines \cite{callahan2016, truong2015}, yielding various data products, one of which are taxonomic tables which can be used to study associations between members of the microbiome and an exposure or outcome of interest.

However, there exists unique challenges in the analysis of these taxonomic count tables \cite{li2019,li2015}. First, like other sequencing-based datasets, microbiome count data is often high dimensional, where the number of detected taxa far exceeds the number of samples usually present. For predictive tasks, microbiome-specific penalized regression approaches have been developed to address this issue \cite{shi2016}, allowing for simultaneous model fitting and variable selection. For differential abundance tasks, researchers often utilize multiple hypothesis correction methods \cite{sankaran2014,benjamini1995} or omnibus tests \cite{chen2018} to address hypothesis testing burden.
